\documentclass[conference]{IEEEtran}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{url}
\usepackage{circuitikz}
\usepackage{booktabs}
\usepackage{array}
\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{green!40!black},
  stringstyle=\color{red},
  numbers=none,                     % no numbers
  numberstyle=\scriptsize\ttfamily, % more readable numbers
  numbersep=8pt,                    % add spacing from code
  xleftmargin=1.5em,                % indent code so numbers don’t overlap
  frame=single,
  breaklines=true,
  breakatwhitespace=true,
  tabsize=2,
  showstringspaces=false,
  captionpos=b,
  keepspaces=true                   % preserve spacing properly
}

\title{Implementation of Cross-Platform Data Transmission Systems: A Comparative Study of Wired Serial and Bluetooth Low Energy (BLE) Bridging using nRF5340}

\author{
\IEEEauthorblockN{Bhargav K, Harsha B J}
\IEEEauthorblockA{
Department of Electrical Engineering\\
IIT Hyderabad\\
Email: ee25btech11013@iith.ac.in, ee25btech11026@iith.ac.in}
}

\begin{document}

\maketitle
\begin{abstract}
This paper presents the design and implementation of a cross-platform bidirectional data transfer system, enabling seamless telemetry and file exchange between a Linux host and an Android mobile environment. The study contrasts a baseline wired serial connection against a wireless embedded bridge utilizing the Nordic Semiconductor nRF5340 SoC.

In the wireless implementation, the nRF5340 functions as a protocol converter, bridging USB-CDC (ACM) serial data to Bluetooth Low Energy (BLE) packets via the Nordic UART Service (NUS). To mitigate data corruption caused by control character interpretation in the UART stack, a Base64 encoding scheme was integrated, successfully enabling the transmission of binary assets (images, PDFs) with 100\% data integrity.
\end{abstract}
\section{Hardware Requirements}
\begin{itemize}
    \item Development Board: Nordic Semiconductor nRF5340 DK.
    \item Smartphone: Android device with Termux, nRF Connect for Mobile and Serial Bluetooth Terminal App.
    \item PC: Linux machine
    
\end{itemize}

\section{Software and Toolchains Installations}
\subsection{1. Update \& Install System Dependencies}
First, update the package manager and install the core dependencies required by the Zephyr build system (CMake, Ninja, Python, etc.).

\begin{lstlisting}[language=bash]
sudo apt update
sudo apt upgrade -y
sudo apt install --no-install-recommends -y \
    git cmake ninja-build gperf \
    ccache dfu-util device-tree-compiler wget \
    python3-dev python3-pip python3-setuptools python3-tk python3-wheel xz-utils file \
    make gcc gcc-multilib g++-multilib libsdl2-dev
\end{lstlisting}

\subsection{2. Install West (Zephyr Build Tool)}
Install \texttt{west}, the command-line tool used to manage the project and build system.

\begin{lstlisting}[language=bash]
pip3 install --user -U west
# Ensure ~/.local/bin is in your PATH if west is not found
export PATH=$PATH:~/.local/bin
\end{lstlisting}

\subsection{3. Initialize nRF Connect SDK (v3.2.2)}
Create a workspace directory and initialize the specific SDK version used for this project.

\begin{lstlisting}[language=bash]
# 1. Create a folder for the SDK
mkdir ~/ncs
cd ~/ncs

# 2. Initialize the SDK pointing to version v3.2.2
west init -m https://github.com/nrfconnect/sdk-nrf --mr v3.2.2

# 3. Download the code
west update

# 4. Export CMake package for Zephyr
west zephyr-export
\end{lstlisting}

\subsection{4. Install Python Dependencies}
Install the specific Python requirements for the Zephyr kernel, the nRF SDK, and the bootloader.

\begin{lstlisting}[language=bash]
pip3 install --user -r zephyr/scripts/requirements.txt
pip3 install --user -r nrf/scripts/requirements.txt
pip3 install --user -r bootloader/mcuboot/scripts/requirements.txt
\end{lstlisting}

\subsection{5. Install the Zephyr SDK Toolchain}
Download and install the cross-compiler toolchain (Zephyr SDK 0.16.1) required to compile code for the nRF5340.

\begin{lstlisting}[language=bash]
# 1. Download the Zephyr SDK Bundle
cd ~
wget https://github.com/zephyrproject-rtos/sdk-ng/
releases/download/v0.16.1/zephyr-sdk-0.16.1_linux-x86_64.tar.xz

# 2. Extract the archive
tar xvf zephyr-sdk-0.16.1_linux-x86_64.tar.xz

# 3. Run the setup script
cd zephyr-sdk-0.16.1
./setup.sh
\end{lstlisting}

\subsection{6. Install nRF Command Line Tools}
Finally, install the J-Link drivers and \texttt{nrfjprog} to enable flashing the board over USB.

\begin{lstlisting}[language=bash]
# Navigate to the download directory
cd ~/Downloads

# Install the .deb package (replace filename with actual version)
sudo dpkg -i nrf-command-line-tools*.deb

# Fix any missing dependencies
sudo apt install -f
\end{lstlisting}

\subsection{Verification}
Verify the installation by checking the versions of the flashing and build tools.

\begin{lstlisting}[language=bash]
nrfjprog --version
west --version
\end{lstlisting}

\section{Project Execution Procedure}

This section details the procedure to compile, flash, and test the \texttt{peripheral\_uart} sample application using the command line interface.

\subsection{1. Workspace Initialization}
Before building, ensure the workspace is correctly set up. If you have not created a specific project folder yet, navigate to your NCS root directory.

\begin{lstlisting}[language=bash]
cd ~/ncs
# Ensure environment variables are set (optional but recommended)
source zephyr/zephyr-env.sh
\end{lstlisting}

\subsection{2. Building the Firmware}
Use the \texttt{west build} command to compile the sample code. We target the \texttt{peripheral\_uart} sample located in the Bluetooth folder.

\textbf{Target Board:} \texttt{nrf5340dk/nrf5340/cpuapp} (Application Core).

\begin{lstlisting}[language=bash]
# Navigate to the main folder
cd ~/ncs

# Build the peripheral_uart sample
west build -b nrf5340dk/nrf5340/cpuapp nrf/samples/bluetooth/peripheral_uart
\end{lstlisting}

\noindent \textbf{Note:} If the build fails due to cached configurations, append the pristine flag \texttt{-p} to clean the build directory:
\begin{lstlisting}[language=bash]
west build -p -b nrf5340dk/nrf5340/cpuapp nrf/samples/bluetooth/peripheral_uart
\end{lstlisting}

\subsection{3. Flashing the Board}
Once the build completes successfully, flash the binary to the nRF5340 development kit.

\begin{enumerate}
    \item Connect the nRF5340 DK to the PC via thImplementation of Cross-Platform Data Transmission Systems: A Comparative Study of Wired Serial and Bluetooth Low Energy (BLE) Bridging using nRF5340e \textbf{IMCU USB} port (Interface MCU).
    \item Run the flash command:
\end{enumerate}

\begin{lstlisting}[language=bash]
west flash
\end{lstlisting}

\noindent If successful, the terminal will display \texttt{Flashing completed}.

\subsection{4. Testing Data Transfer}
To verify functionality, we establish a connection between the PC terminal and a smartphone.

\subsubsection{A. PC Setup (Receiver)}
Open a serial terminal to view data sent from the phone.
\begin{itemize}
    \item \textbf{Command:} Use \texttt{screen} or \texttt{minicom} to listen to the board's serial output.
    \item \textbf{Baud Rate:} 115200.
\end{itemize}

\begin{lstlisting}[language=bash]
# List available USB devices to find the port (e.g., /dev/ttyACM0)
ls /dev/ttyACM*

# Open the serial connection
sudo screen /dev/ttyACM0 115200
\end{lstlisting}

\subsubsection{B. Smartphone Setup (Sender)}
\begin{enumerate}
    \item Open the \textbf{nRF Connect for Mobile} app on Android.
    \item Scan for devices and locate \textbf{Nordic\_UART\_Service}.
    \item Tap \textbf{Connect}.
    \item Navigate to the \textbf{UART} tab.
    \item Type a text message (e.g., ``Hello Linux'') and send it.
\end{enumerate}

\subsubsection{C. Result}
The text typed on the smartphone should immediately appear in the Linux terminal window. Conversely, typing in the Linux terminal will send text to the smartphone app logs.



\section{File Transfer : Linux Terminal \& Android Serial App}

This section details the procedure to verify bidirectional text transfer using a Linux terminal on the PC side and the \textbf{Serial Bluetooth Terminal} app on an Android device.

\subsection{1. Linux Setup (Receiver/Sender)}
On the Ubuntu machine, we interact with the nRF5340 board through its virtual serial port.

\begin{enumerate}
    \item Connect the nRF5340 DK to the PC via the \textbf{IMCU USB} port.
    \item Open a terminal window and list the USB devices to identify the board (usually \texttt{ttyACM0} or \texttt{ttyACM1}).
\begin{lstlisting}[language=bash]
ls /dev/ttyACM*
\end{lstlisting}
    \item Open the serial connection using \texttt{screen} or \texttt{minicom} with a baud rate of \textbf{115200}.
\begin{lstlisting}[language=bash]
# Replace ttyACM0 with your actual port
sudo screen /dev/ttyACM0 115200
\end{lstlisting}
    \item The terminal screen will clear and wait for input. (Press \textbf{Reset} on the board if you want to see the boot log).
\end{enumerate}

\subsection{2. Android Setup (Serial Bluetooth Terminal)}
We use a dedicated terminal app to act as the Bluetooth Central device.

\begin{enumerate}
    \item \textbf{Install App:} Download and install \textbf{Serial Bluetooth Terminal} (by Kai Morich) from the Google Play Store.
    \item \textbf{Open App:} Launch the application.
    \item \textbf{Scan:}
    \begin{itemize}
        \item Tap the menu (three lines) in the top-left corner.
        \item Select \textbf{Devices}.
        \item Tap the \textbf{Bluetooth LE} tab (important: do not use the Classic Bluetooth tab).
        \item Tap \textbf{Scan} at the top.
    \end{itemize}
    \item \textbf{Connect:} Locate the device named \textbf{Nordic\_UART\_Service} in the list and tap it to connect. The terminal screen should show ``Connected''.
\end{enumerate}

\subsection{3. Verifying File Transfer (Text Streaming)}

This section details the procedure to transfer entire text files between the Android device and the Linux machine using the Nordic UART Service.

\subsubsection{Smartphone to Linux (Sending a File)}
To send a text file stored on the phone to the Linux terminal:
\begin{enumerate}
    \item Ensure the devices are connected via the \textbf{Serial Bluetooth Terminal} app.
    \item Tap the \textbf{Menu} icon (three lines) in the top-left corner or the \textbf{Upload/Folder} icon (if available on the toolbar).
    \item Select \textbf{Send File}.
    \item Browse and select a \texttt{.txt} file from your phone's storage.
    \item \textbf{Result:} The content of the file will stream immediately into the open \texttt{screen} session on your Linux terminal. This can be saved as a file in the PC.
\end{enumerate}

\subsubsection{Linux to Smartphone (Sending a File)}
To send a text file from the Linux machine to the phone, we cannot just type in the terminal. We must pipe the file directly to the device port.

\begin{enumerate}
    \item Open a \textbf{new} terminal window (keep the \texttt{screen} session open in the first window to monitor the connection).
    \item Create a dummy text file to send:
\begin{lstlisting}[language=bash]
echo "This is a test file sent from Linux." > test_file.txt
\end{lstlisting}
    \item Use the \texttt{cat} command to redirect the file into the USB serial port:
\begin{lstlisting}[language=bash]
# Replace ttyACM0 with your actual port
sudo sh -c 'cat test_file.txt > /dev/ttyACM0'
\end{lstlisting}
    \item \textbf{Result:} The content of \texttt{test\_file.txt} will appear in the \textbf{Serial Bluetooth Terminal} app on your phone (usually in blue).
\end{enumerate}

\section{Transferring Binary Files (Images/PDFs) via Base64}
The Nordic UART Service (NUS) is designed for character-based data streams. Attempting to send raw binary files (e.g., \texttt{.jpg}, \texttt{.png}, \texttt{.pdf}) directly will result in data corruption due to control characters and null bytes. To transfer these files, we must encode them into a text-safe format using Base64.

\subsection{1. Concept Overview}
The process involves three stages:
\begin{enumerate}
    \item \textbf{Sender (Linux):} Encodes the binary file into a Base64 text string.
    \item \textbf{Transmission:} The text string is sent over the USB-to-Bluetooth bridge.
    \item \textbf{Receiver (Termux):} Receives the text string and decodes it back into the original binary format.
\end{enumerate}

\subsection{2. Procedure: Linux to Android (Termux)}

\subsubsection{Step A: Prepare the Receiver (Termux)}
On the Android device, open Termux and set up a listener that will dump the incoming data into a temporary file.

\begin{lstlisting}[language=bash]
# 1. Open a listening connection (via the Bridge App on port 9000)
# 2. Redirect output to a file named 'received_data.b64'
nc 127.0.0.1 9000 > received_data.b64
\end{lstlisting}
\textit{Termux is now waiting for data...}

\subsubsection{Step B: Send the File from Linux}
On the Linux laptop, we use the \texttt{base64} command to encode the file and pipe it immediately to the USB port.

\begin{lstlisting}[language=bash]
# Syntax: base64 <input_file> > <device_port>

# Example: Sending an image named 'photo.jpg'
base64 photo.jpg > /dev/ttyACM0
\end{lstlisting}
\textit{The transfer will take a few seconds depending on the file size. Once the command finishes on Linux, close the connection on Termux (Ctrl+C).}

\subsubsection{Step C: Decode the File on Termux}
Now that we have the Base64 text file on the phone, we must decode it back to its original binary format.

\begin{lstlisting}[language=bash]
# 1. Decode the text file back to binary
base64 -d received_data.b64 > restored_photo.jpg

# 2. Verify the file exists and has size
ls -lh restored_photo.jpg
\end{lstlisting}

\subsection{3. Procedure: Android (Termux) to Linux}

\subsubsection{Step A: Prepare the Receiver (Linux)}
On the Linux laptop, open a terminal to listen for the incoming stream.

\begin{lstlisting}[language=bash]
# Read from the USB port and save to a file
cat /dev/ttyACM0 > received_from_phone.b64
\end{lstlisting}

\subsubsection{Step B: Send the File from Termux}
In Termux, locate the file you want to send and pipe it to the bridge connection.

\begin{lstlisting}[language=bash]
# Encode a PDF file and send it to the bridge port (9000)
base64 document.pdf | nc 127.0.0.1 9000
\end{lstlisting}

\subsubsection{Step C: Decode on Linux}
Once the transfer stops, press \textbf{Ctrl+C} on the Linux terminal to stop recording. Then decode the file.

\begin{lstlisting}[language=bash]
# Decode back to PDF
base64 -d received_from_phone.b64 > restored_document.pdf

# Open the file to verify
xdg-open restored_document.pdf
\end{lstlisting}

\subsection{4. Technical Constraints}
\begin{itemize}
    \item \textbf{Overhead:} Base64 encoding increases the file size by approximately 33\%.
    \item \textbf{Speed:} Bluetooth Low Energy (BLE) has a limited bandwidth (approx. 10-20 kB/s for NUS). Transferring large files (e.g., $>$ 1MB) may take several minutes.
    \item \textbf{Buffer:} Ensure the receiving device does not sleep during transfer, as BLE connections may drop if the radio is idle for too long.
\end{itemize}


\section{File Transfer between 2 devices via bluetooth without using nRF5340 BLE}
Switch on bluetooth on both devices.
\subsection{Installating the Packages}
\begin{lstlisting}[language=bash]
sudo apt update
sudo apt install bluez bluetooth bluetooth-obex obexftp gnome-bluetooth
sudo systemctl start bluetooth
\end{lstlisting}
\subsection{Connecting the sender to the receiver}
On the sender device:
\begin{lstlisting}[language=bash]
bluetoothctl scan on
\end{lstlisting}
Note the MAC address: AA:BB:CC:DD:EE:FF
\begin{lstlisting}[language=bash]
bluetoothctl
pair AA:BB:CC:DD:EE:FF
trust AA:BB:CC:DD:EE:FF
connect AA:BB:CC:DD:EE:FF
exit
\end{lstlisting}

\begin{lstlisting}[language=bash]
bluetooth-sendto --device=AA:BB:CC:DD:EE:FF /path/to/file1.txt /path/to/file2.jpg /path/to/file3.png
\end{lstlisting}

To automate this process: \\
Save the following as 
bt-multi-send.sh
\begin{lstlisting}[language=bash]
#!/bin/bash
DEVICE="AA:BB:CC:DD:EE:FF"  # Replace with MAC Address
DIR_PATH="$1"  # Pass dir as arg, e.g., ./bt-multi-send.sh /path/to/files

if [ -z "$DIR_PATH" ]; then
  echo "Usage: $0 <directory>"
  exit 1
fi

# Ensure paired
bluetoothctl connect $DEVICE > /dev/null 2>&1

for file in "$DIR_PATH"/*; do
  if [ -f "$file" ]; then
    echo "Sending $file..."
    bluetooth-sendto --device=$DEVICE "$file"
  fi
done

\end{lstlisting}

Make it executable:
\begin{lstlisting}[language=bash]
chmod +x bt-multi-send.sh
\end{lstlisting}

Run this command:
\begin{lstlisting}[language=bash]
./bt-multi-send.sh /path/to/files
\end{lstlisting}

\section{Industry Applications of Nordic SoC's}

Nordic Semiconductor SoCs are widely adopted in commercial IoT sectors. Notable examples include:

\begin{itemize}
    \item \textbf{Wearables:} The \textit{Ultrahuman Ring Air} and \textit{Dayton Smartwatch} utilize the nRF52 series for biometric tracking (ECG, Sleep Index).
    \item \textbf{Medical Healthcare:} Devices such as \textit{PulseTape} utilize Nordic chips for continuous ECG monitoring, prioritizing low power consumption for extended battery life.
    \item \textbf{Smart Home:} Significantly, the \textit{eCozy 2.0 Smart Thermostat} employs the \textbf{nRF5340}—the exact SoC utilized in this project—to handle complex connectivity protocols like Matter and Bluetooth Mesh.
\end{itemize}
\section{Conclusion}

This project successfully demonstrated the implementation of a robust, bidirectional wireless data transfer system utilizing the Nordic Semiconductor nRF5340 System-on-Chip. By configuring the nRF5340 as a bridge between a USB-CDC (Serial) interface and the Bluetooth Low Energy (BLE) Nordic UART Service (NUS), we established a seamless communication link between a Linux-based PC and an Android mobile environment and between 2 Linux-based PC.


\begin{itemize}
    \item \textbf{Firmware Implementation:} The firmware was successfully built and flashed using the Zephyr RTOS command-line toolchain (\texttt{west}), demonstrating a professional, reproducible development workflow without reliance on heavy GUI IDEs.
    \item \textbf{Data Integrity:} While the UART protocol is inherently text-based, we successfully implemented a Base64 encoding strategy. This enabled the error-free transmission of complex binary files (images, PDFs) over the link, proving the system's versatility beyond simple telemetry.
\end{itemize}

\subsection{Final Verdict}
The comparative analysis between direct wired connections and the nRF5340-based solution highlights the flexibility of BLE for embedded applications. While wired connections offer higher raw throughput, the nRF5340 implementation provides a reliable, low-power wireless alternative suitable for IoT remote debugging, field telemetry, and file exchange in environments where physical cabling is impractical. The system stands as a functional proof-of-concept for modern embedded wireless bridging.


\end{document}

